Finch: A Datastructure-Driven Array Programming Language

### 2024-03-18 main.tex
Update on Overleaf.

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{amssymb}
\usepackage{array}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

\author{Teo Collins}
\affiliation{%
  \institution{MIT CSAIL}
  \city{Cambridge}
  \state{Massachusetts}
  \country{USA}}
\email{teoc@mit.edu}

@@

%- Fortran supported lots of control structures, no data structures, just dense arrays.
%
%- We tried to emulate complex data using only dense arrays but with complicated control structures.
%
%- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
%
%- We’re bringing these together, both need to work together.
\subsection{Contributions}

\begin{enumerate}
\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. To our knowledge, the Finch programming language is the first
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather.
\item More complex array structures than ever before. A complete level-by-level
structure-description language for expressing the structure of data
hierarchically. The first such set of formats to efficiently capture banded,
triangular, run-length-encoded, or sparse datasets, and any combination thereof.
\item The Finch compiler specializes programs to data structures in a
predictable, deterministic approach, making it easier to search the complex
space of programs and datastructures to find an appropriate fit for a given
application. The tensor interface makes it easy to extend Finch to new level
formats. A unique tensor lifecycle model enables polymorphism by analyzing the
appropriate stages to insert simple, overloadable, interface functions such as
initialization or finalization.

\item We evaluate the productivity of our language in several case studies,
showing that Finch can be used to accelerate a wide range of applications, 
from classic operations such as spmv and spgemm, to more complex applications such as image processing and graph analytics.
We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for
operations such as map, broadcast, or reduce that can be compiled to efficient
code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

@@

@@

\subsubsection{SparseList Format}
\subsubsection{SparseVBL Format}
The SparseVBL format is similar to the SparseList format, but contiguous fibers are stored together as blocks. Thus, it is most worthwhile to use when the matrix has a blocked structure—i.e. where there tend to be stretches of consecutive nonzero values. We found that the SparseVBL format performed the best for matrices with a clear block structure (list specific matrices) or an imperfect blocked structure (list specific matrices). 
\subsubsection{Pattern Format}
\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO . Notably, there is a 1.9x speedup for the HB/saylr4 matrix. 
\subsubsection{4D Blocked SpMV}

@@

@@

\begin{figure}
	\includegraphics[width=\linewidth]{fill_speedup_over_opencv.png}
    \caption{Performance of Finch on flood fill task.}
\end{figure}

@@

>>>>

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{amssymb}
\usepackage{array}
\usepackage{minted}

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

\author{Teodoro Fields Collin}
\affiliation{%
  \institution{MIT CSAIL}
  \city{Cambridge}
  \state{Massachusetts}
  \country{USA}}
\email{teoc@mit.edu}

@@

%- Fortran supported lots of control structures, no data structures, just dense arrays.
%
%- We tried to emulate complex data using only dense arrays but with complicated control structures.
%
%- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
%
%- We’re bringing these together, because if both of these can work together, we can more efficiently search the space of programs

\textbf{Finch makes high-performance computing easier with compilers that adapt programs to the structure of data}.  Currently, experts must hand write variations of even the simplest kernels, like matrix multiply, for each data structure and architecture to get performance. This makes performance engineering an expensive, repetitive task. Several popular programs compute the same thing on different data.

\textbf{Unstructured data is the easiest way to program, but the world is full of structured data}. Multidimensional dense (unstructured) arrays are the most important abstraction in computer science. Since their introduction in the 1950s with Fortran and Lisp, the simplicity of dense arrays has enabled extensive innovation, from compiler optimizations to convenience layers. However, structure is everywhere. Sparse arrays (which store only nonzero elements) can describe networks and simulations. Run-length encoding describes images, geometry and databases (such as a list of transactions with the date field all the same). Symmetry, bands, padding, and blocks are all natural properties which arise in the course of scientific computing, simulating the world around us.

\textbf{Structured data support is fragmented and incomplete}. Programmers today must choose between program flexibility and datastructure flexibility. Hand-written solutions are collected in diverse libraries like MKL, OpenCV, or SciPy. However, libraries will only ever support a subset of programs on a subset of datastructure combinations. Special-purpose compilers like TACO or CoRa support a select subset of data structures (only sparse, or only ragged arrays) in highly constrained classes of programs such as tensor contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}.

\textbf{Finch handles the two main challenges for structured array data}:
\begin{itemize}
\item \textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory. For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value. These zero regions or repeated regions are optimization opportunities. Here, we show how Finch combines programs with datastructure descriptions to generate efficient, datastructure-specific code.

\item \textbf{Performance is highly structure dependent: Structure aware kernels are dramatically faster than unstructured ones, but the landscape of implementation decisions is also dramatically unpredictable}. For example, sparse kernels don’t need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation. Finch allows you to search it more efficiently.
\end{itemize}

\subsection{Finch: Compiling Structured Data}
The finch array programming language adapts complex programs to array datastructures to produce efficient code. Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). Finch supports \textit{both} flexible programming constructs and diverse data structures. Finch supports loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.

Crucially, Finch is able to adapt simple, unstructured programs to the structure of the input data. It does this using Looplets, a new language that decomposes datastructure iterators hierarchically. Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program. Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation.

\subsection{Contributions}

\begin{enumerate}
\item More complex array structures than ever before. A complete level-by-level
structure-description language for expressing the structure of data
hierarchically. The first such set of formats to efficiently capture banded,
triangular, run-length-encoded, or sparse datasets, and any combination thereof.
\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. To our knowledge, the Finch programming language is the first
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather.
\item A compiler that specializes programs to data structures 
predictably, facilitating an expressive language that makes it easier to search the complex
space of algorithms and datastructures. It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
initialization or finalization.
\item We evaluate the productivity of our language in several case studies,
showing that Finch can be used to accelerate a wide range of applications, 
from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension. Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense. Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
\end{figure}

@@

% TODO: make this be 3 columns
\begin{figure}
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[i] += A[i, j] * x[j]
                end
                return y
            end
        \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[j] += A[i, j] * x[i]
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \begin{minipage}[t]{0.36\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _
                    let x_j = x[j]
                        y_j .= 0
                        for i = _
                            let A_ij = A[i, j]
                                y[i] += x_j * A_ij
                                y_j[] += A_ij * x[i]
                            end
                        end
                        y[j] += y_j[] + diag[j] * x_j
                    end
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
\end{figure}

\subsubsection{Tensor Formats}
We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix. To represent scalar values, we benchmarked with both the Element and Pattern levels. The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices. The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.

We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor. Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation. Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.

There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean. The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.

\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO . Notably, there is a 1.9x speedup for the HB/saylr4 matrix. 
\subsubsection{4D Blocked SpMV}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (2 iterations).}
\end{figure}

@@

@@

Push-Pull BFS in Finch:
\begin{minted}{julia}
V = Tensor(Dense(Element(false)))
P = Tensor(Dense(Element(0)))
F = Tensor(SparseByteMap(Pattern()))
_F = Tensor(SparseByteMap(Pattern()))
A = Tensor(Dense(SparseList(Pattern())))
AT = Tensor(Dense(SparseList(Pattern())))

function finch_bfs_push_kernel(_F, F, A, V, P)
    @finch begin
        _F .= false
        for j=_, k=_
            if F[j] && A[k, j] && !(V[k])
                _F[k] |= true
                P[k] <<choose(0)>>= j #Only set the parent for this vertex
            end
        end
        return _F
    end
end

function finch_bfs_pull_kernel(_F, F, AT, V, P)
    p = ShortCircuitScalar{0}()
    @finch begin
        _F .= false
        for k=_
            if !V[k]
                p .= 0
                for j=_
                    if F[follow(j)] && AT[j, k]
                        p[] <<choose(0)>>= j #Only set the parent for this vertex
                    end
                end
                if p[] != 0
                    _F[k] |= true
                    P[k] = p[]
                end
            end
        end
        return _F
    end
end)
\end{minted}

### 2024-03-18 sections/tensors.tex
Update on Overleaf.

Looplets provide descriptions of iterators. 
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.

%
As we will see
\begin{enumerate}
    \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
    \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
    \item how do we ask for the right iterator and how do we change/manage their state?
\end{enumerate}
Need to Make these problems clear by this point, hope that read anticipate them.

Solutions:
\begin{enumerate}
    \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
    \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
\end{enumerate}

@@

\subsection{Core Level Langage Primitives}
\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}

>>>>

Looplets provide descriptions of iterators. 
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
%
First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is much easier to manage when considering a single dimensional iterator. 
%
We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
%
Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
%
So we require a bridge that can provide different iteration orders depending on the context.
%
Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.

To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
%
Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}. 
%
We recapitulate the essential details here.
%
Our second abstraction adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation and limiting the amount of specific work.
%
Previous efforts to compiler a greater variety of sparse tensor programs left this bridge untouched ~\cite{henry2021compilation, won2023unified}.
%
Our third abstraction limits our set of sparse tensor programs to accommodate a natural restriction on when multi-dimensional data structures can be modified.
%
% As we will see
% \begin{enumerate}
%     \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
%     \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
%     \item how do we ask for the right iterator and how do we change/manage their state?
% \end{enumerate}
% Need to Make these problems clear by this point, hope that read anticipate them.

% Solutions:
% \begin{enumerate}
%     \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
%     \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
% \end{enumerate}

@@

\subsection{Instatiation Interface}

\subsection{Core Level Language Primitives}
\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}
